---
layout: post
title: "Reordering Bar Charts with ggplot"
modified:
categories: r
excerpt:  "A simple way to reorder factors to use in bar charts + ggplot."
tags: [R, ggplot, data visualisation]
comments: true
share: true
image:
  feature:
date: 2016-04-18
---
  
Found this solution on stack overflow and it was brilliant!
  
[http://stackoverflow.com/questions/5967593/ordering-of-bars-in-ggplot](http://stackoverflow.com/questions/5967593/ordering-of-bars-in-ggplot)

In particular, I'd never heard of the `reorder()` command before and would 
usually reorder the data frame for plotting manually.

Here's how I would have normally done it:
  
```{r}
## Notice the as.is = TRUE
breadth_data <- read.table(textConnection("Stakeholder  Value
                                          'Grantseekers'  0.90
                                          'Donors'    0.89
                                          'Community' 0.55
                                          'Hurricane Relief Fund' 0.24
                                          'Media' 0.19
                                          'Employment Seekers'    0.12
                                          'Affiliates'    0.10
                                          'Youth' 0.09
                                          'Women' 0.02
                                          'Former Board Members'  0.01"), 
                           header=TRUE, 
                           as.is = TRUE)
str(breadth_data)
```

Since the `Stakeholder` column in the data frame are factors, you can specify 
the factors order for this data frame. Here, I've ordered it in ascending order 
of `Value`.

```{r}
# reorder manually
orderIdx <- order(breadth_data$Value)
breadth_data <- breadth_data[orderIdx, ]
breadth_data$Stakeholder <- factor(breadth_data$Stakeholder,
                                   levels = breadth_data$Stakeholder,
                                   labels = breadth_data$Stakeholder)
levels(breadth_data$Stakeholder)

# plot
library(ggplot2)
ggplot(breadth_data, aes(x = Stakeholder, y = Value)) + 
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_bw()
```

With the `reorder()` function, this whole process is simplified.  Let's start the whole 
process again:

```{r}
breadth_data.new <- read.table(textConnection("Stakeholder  Value
                                          'Grantseekers'  0.90
                                              'Donors'    0.89
                                              'Community' 0.55
                                              'Hurricane Relief Fund' 0.24
                                              'Media' 0.19
                                              'Employment Seekers'    0.12
                                              'Affiliates'    0.10
                                              'Youth' 0.09
                                              'Women' 0.02
                                              'Former Board Members'  0.01"), 
                               header=TRUE)
breadth_data.new <- transform(breadth_data.new,
                          Stakeholder = reorder(Stakeholder, Value))

ggplot(breadth_data.new, aes(x = Stakeholder, y = Value)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_bw()


```

